### 2012-05-27 trunk/ijcis-pos-model/sources/temporaldatabases.tex
modifications on the operations

\begin{definition}
Consider a relation $R$ in a temporal database, an object $A$ given by the attributes $\left(a_1, \ldots, a_n \right)$ and the crisp time interval $I$ given by $\left[a,\ b\right] : a,b \in \T, a \leq b$. A valid-time object $A$ in the relation $R$ is specified by:
\begin{equation}
\label{eq:rel-def}
\left( R, A , I \right)
\end{equation}
\end{definition}

In order to simplify the algorithms for the manipulation of data, some auxiliary functions are defined:

\begin{definition}
\label{def:close-a-crisp-interval-r}
Consider two crisp intervals $I= \left[a ,b \right]$ and $J= \left[c ,d \right]$, $c > a$. The CloseR$\left(I, J\right)$ function allows to close the right-open interval $I$ with respect to the first value $d$ in $J$:
\begin{align}
\label{eq:close-a-crisp-interval}
\mbox{CloseR} \left( I, J \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } b \neq +\infty \\
I=\left[a, c \right[ & \mbox{ if } b = +\infty \wedge J > I
\end{cases}
\end{align}
\end{definition}

@@

\begin{definition}
\label{def:current-in-relation}
Consider an object $A$ and a relation $R$ in a temporal database. The function Current$\left(R, A \right)$ obtains the tuple $\left(A, I \right)$ that is current in the relation $R$, that is it, $I = \left[a, +\infty \right]$.

\begin{align}
\label{eq:current-in-relation}
\mbox{Current} \left(R, A \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } \exists b \in I : (A,I) \in R \wedge b = +\infty \\
\emptyset & \mbox{ in any other case }
\end{cases}
\end{align}
\end{definition}

@@

Now it is possible to close the current version of an object by using \eqref{def:close-a-crisp-interval-r} and \eqref{eq:current-in-relation}

\begin{definition}
\label{def:close-current-version}
Consider an object $A$, the relation $R$ and a time interval $J$. Then, the function close-current$\left(R,A,J \right)$ closes any current version of the object $A$ if it exists and add the new version $\left(A, J \right)$.

\begin{eqnarray}
\label{eq:close-current}
\text{Close-current} \left(R, A, J \right) =\\
\begin{cases}
\nonumber
\big \lbrace R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseR } \left(I, J\right) \right) \cup \left(A, J\right)\right \rbrace  \big \rbrace \\
\nonumber
\mbox{ if } I = \mbox{ Current } \left(R, A \right) \wedge J > \mbox{ CloseR } \left(I, J \right)   \\
\nonumber R , \text{ in any other case}
\end{cases}
\end{eqnarray}
\end{definition}

%It is also possible to append at the beggining by using \eqref{def:close-a-crisp-interval-l} and \eqref{def:first-in-relation}.
%\begin{definition}
%\label{def:append-first-version}
%Consider an object $A$, the relation $R$ and a time interval $J$. Then, the function append-first$\left(R,A,J \right)$ closes any right open version of the object $A$ if it exists and add the new version $\left(A, J \right)$.
%
%\begin{eqnarray}
%%\label{eq:close-current}
%\text{Append-first} \left(R, A, J \right) =\\
%\begin{cases}
%\nonumber
%\big \lbrace R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseL } \left(I, J\right) \right) \cup \left(A, J\right)\right \rbrace  \big \rbrace \\
%\nonumber
%\mbox{ if } I = \mbox{ First } \left(R, A \right) \wedge J < \mbox{ CloseL } \left(I, J \right)   \\
%\nonumber R , \text{ in any other case}
%\end{cases}
%\end{eqnarray}
%\end{definition}

\subsubsection{\label{subsubsec:insert}Insert}
The user wants to store an object $A$ which is valid in the relation $R$ during the time interval specified by $I = \left(a, b \right)$.
%
%
%The interpretation of the insert operation is the following: The user wants to store in the database that an object $A$  is true for some period(s) of time $I$ in the given relation $R$. To indicate that the object $A$ is current in the relation, the special value Until Current, $\UC$  is used. 
There are two main cases when performing a create operation:
\begin{enumerate}
\item The object $A$ was never in the relation $R$: The object is added with the valid-time indicated by the crisp interval $I$.

\item The object $A$ is in the relation $R$. Depending on the value of the time interval $I$, there are two possibilities:
	\begin{enumerate}
	\item Insert $\left(A, I\right)$ in the relation $R$. If the time interval $I$ does not overlap any other valid-time interval in the relation $R$.
	\item Reject the insertion, if the time interval $I$ do overlap any existing valid-time interval for the object $A$ in the relation $R$.
	\end{enumerate}

@@

@@

\begin{align}
\label{eq:insert}
\INS \left(R ,A, I\right) =\\
\begin{cases}
\nonumber
R \cup \left(A, I \right), & \mbox{ if }  A \not \in R\\
R \cup   \left(A, I \right) & \mbox{ if }  A \in R \wedge \\
& \big \lbrace \left( \exists J: \left(A, J \right) \in R \wedge J<I  \right) \vee \\
& \left( \exists K: \left(A, K \right) \in R \wedge K>I \vee \right) \wedge\\
&  J<I<K  \big \rbrace\\
R, & \mbox{ otherwise }  
\end{cases} 	
\end{align}

\subsubsection{\label{subsubsec:del}Delete}
The delete operation logically removes a current object $A$ which is valid in the relation $R$:
\begin{align}
\label{eq:delete}
\DEL \left(R ,A \right) =\\
\begin{cases}
\nonumber
R - \left(A, I \right), & \mbox{ if } \left( A, I \right) \in R\\
R, & \mbox{ otherwise }  
\end{cases} 	
\end{align}

\subsubsection{\label{subsubsec:modify}Modify}
This operation adds new information about an existing object $A$ in the relation $R$. The modify operation does not remove any previous value of the object $A$.
Note that the modify operation is only applicable when the object $A$ is current in the relation $R$; $ \left(A, I \right) \in R, I = \left[a, +\infty \right]$.

\begin{align}
\label{eq:modify}
\MOD \left( R, A, I \right) =\\
\begin{cases}
\nonumber
\mbox{ Close-current }\left(R, A, \mbox{Current}\left(R, A \right) \right) \\
 \mbox{ if Current}\left(R ,A \right) \neq \emptyset \\
 R,  \mbox{ otherwise }
\end{cases}
\end{align}

\subsubsection{\label{subsubsec:revise}Revise}
This operation changes the value of the object $A$ in the relation $R$, at a given time interval $I$. This operation is defined when a correction on the values for the object $A$ should be done.
\begin{align}
\label{eq:revise}
\mbox{ Revise}\left(R, A, I \right)=\\
\begin{cases}
\nonumber
 R - \left(A', I \right) \cup \left(A, I \right)  \\
  \mbox{ if } \left(A', I \right) \in R: A = \left(a_1, \ldots, a_n \right),\\
A'= \left(a_1', \ldots, a_n' \right)   \\   
R,  \mbox{ otherwise }
\end{cases}
\end{align}

>>>>

\begin{definition}
Consider a relation $R$ in a temporal database, an entity $A$ given by the attributes $\left(a_1, \ldots, a_n \right)$ and the crisp time interval $I$ given by $\left[a,\ b\right] : a,b \in \T, a \leq b$. The pair $\left(A, I\right)$ expresses that the data regarding the entity A is valid during the crisp time interval I.  Note that the relation $R = \left(A, I \right) x \left(A, I \right)$. The following equation indicates that the pair $\left(A, I\right)$ is in the relation R:
\begin{equation}
\label{eq:rel-def}
\left( R, \left(A , I\right) \right)
\end{equation}
\end{definition}
\begin{example}
 Consider a library database. Each physical book is stored with an identifier (ID), a title and the start and end loan dates respectively. Table \ref{tbl:library-sample} illustrates the first version of the database, after three insertions. In this example, $A = ($ ID, Title $)$ and $I = ($ Start Loan, End Loan $)$.
\end{example}

\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample}Sample library database}
\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Title  & Start Loan & End Loan \\
\hline
3 & Dracula & 15/3/2012 & 30/3/2012 \\
4 & Frankenstein & 25/3/2012 & 4/4/2012 \\
5 & Harry Potter & 18/3/2012 & 2/4/2012\\
3 & Dracula & 4/4/2012 & UC \\
\hline\\
\end{tabular}
}

In order to simplify the algorithms for the manipulation of data, some auxiliary functions and constants are defined:

\begin{definition}
 \label{def:until-changed}
\textbf{Until changed (UC):} To show that the data related to the entity $\left(A, I \right)$ is up-to-date, a special constant called \emph{Until Changed} is defined for the endpoint of the time interval $I$. The value for this constant is considered to be $+\infty$. 
\end{definition}

%quizas poner un ejemplo relacionado con lo de la liberia.

If the pair $\left(A, I \right)$ with $I = \left(c, UC \right), c \in \T$ is in the relation $R$, then, it is said that the pair $\left(A, I \right)$ is \emph{current} in the relation.

For example, last row in Table \ref{tbl:library-sample}. The value for the time interval is $I = \left(4/4/2012 , UC \right)$. The meaning is that the book with ID = 3 was loaned the  4/4/2012 and it is still loaned. The book with ID = 3 is current in the relation.

\begin{definition}
\textbf{Current $\left(R, A \right)$:}
\label{def:current-in-relation}
This function returns a crisp time interval $I$ if the entity $A$ is current in the relation $R$ or $\emptyset$ if the entity $A$ is not current in the relation.

%Consider an object $A$ and a relation $R$ in a temporal database. The function Current$\left(R, A \right)$ obtains the tuple $\left(A, I \right)$ that is current in the relation $R$, that is it, $I = \left[a, +\infty \right]$.

\begin{align}
\label{eq:current-in-relation}
\mbox{Current} \left(R, A \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } (A,I) \in R \wedge I = \left(c, UC \right), c \in \T\\
\emptyset & \mbox{ in any other case }
\end{cases}
\end{align}
\end{definition}

For example, the function Current(R, (3,`Dracula')) returns the time interval $I = \left(4/4/2012 , UC  \right)$. On the other hand, Current(R, (4,`Frankenstein')) returns the empty set.

The following two Allen's relations (see Figure \ref{fig:allen-rel}) are defined like in~\cite{Nagypal2003} to implement the auxiliary functions.

\vspace*{13pt}
\begin{center}
{
\includegraphics[scale=0.5]{./graphs/allen.pdf}

}
\end{center}
%\centerline{ \psfig{file=./graphs/Y-time-point.eps}}
\vspace*{10pt}
\fcaption{\label{fig:allen-rel}Allen's relations for two crisp time intervals $I$ and $J$.}
\vspace*{13pt}

\begin{definition}
\textbf{Overlaps:}
 \label{def:overlaps}
Given two time intervals $I = \left(c, d \right)$ and $J = \left(e, f\right)$, it is said that I overlaps J if:
\begin{equation}
 \label{eqn:overlaps}
\text{I Overlaps J}  = \left(c > e  \right) \wedge \left(d < f  \right) 
\end{equation}
\end{definition}

\begin{definition}
\textbf{During:}
 \label{def:during}
Given two time intervals $I = \left(c, d \right)$ and $J = \left(e, f\right)$, it is said that I during J if:
\begin{equation}
 \label{eqn:during}
\text{I During J}  = \left(c < e  \right) \wedge \left(d > e  \right) \wedge \left(d < f \right)
\end{equation}
\end{definition}

\begin{definition}
\textbf{CloseR$\left( I, J \right)$:}
\label{def:close-a-crisp-interval-r}
Consider two crisp intervals $I= \left[c ,d \right]$ and $J= \left[e ,f \right]$. The CloseR$\left(I, J\right)$ function allows to close the right-open interval $I$ with respect to the first value $e$ in $J$:
\begin{align}
\label{eq:close-a-crisp-interval}
\mbox{CloseR} \left( I, J \right) &=& \\ 
\begin{cases}
\nonumber
%I & \mbox{ if } b \neq +\infty \\
%I=\left[a, c \right[ & \mbox{ if } b = +\infty \wedge J > I
\left[c, e \right[ & \mbox{ if } d = UC \wedge J \mbox{ During } I \\
I & \mbox{ in any other case.}
\end{cases}
\end{align}
\end{definition}

For example, consider two time intervals, $I = \left(4/4/2012, UC \right)$ and $J = \left(24/4/2012, UC \right)$. The result of applying the CloseR$\left(I, J \right)$ is $I = \left(4/4/2012, 23/4/2012 \right)$.

@@

@@

Now it is possible to close the current version of an entity by using \eqref{eq:close-a-crisp-interval} and \eqref{eq:current-in-relation}. This functionality is required to append new information about an existing entity in the relation.

\begin{definition}
\label{def:close-current-version}
Consider an entity $A$, the relation $R$ and a time interval $J$. Then, the function close-current$\left(R,\left(A,J\right) \right)$ closes any current version of the object $A$ if it exists and add the new version $\left(A, J \right)$.

\begin{eqnarray}
\label{eq:close-current}
\text{Close-current} \left(R, \left(A, J\right) \right) =\\
\begin{cases}
\nonumber
R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseR } \left(I, J\right) \right)\right \rbrace \cup \left \lbrace\left(A, J\right) \right \rbrace  \\
\nonumber
\mbox{ if } I = \mbox{ Current } \left(R, A \right)\\ %\wedge J > \mbox{ CloseR } \left(I, J \right)   \\
\nonumber R , \text{ in any other case}
\end{cases}
\end{eqnarray}
\end{definition}

For example, consider the book with ID = 3 in Table \ref{tbl:library-sample}. The current version was loaned on 4/4/2012. The book was returned to the library but, for some reason, the end loan date was not registered. The book is loaned again on 24/4/2012. Then, the Close-current function allows to close the current version of the book and insert a new version. First, the function CloseR is applied with $I = \left(4/4/2012, UC \right)$ and $J = \left(24/4/2012, UC \right)$. Hence, the value for the time interval $I$ is $ \left(4/4/2012, 23/4/2012 \right)$. Then, the modifications on the value for the time interval $I$ are stored. Finally, a new row with the current values of the book and the time interval $J$ is stored.
The result of this operation is illustrated on Table \ref{tbl:library-sample-2}.

%It is also possible to append at the beggining by using \eqref{def:close-a-crisp-interval-l} and \eqref{def:first-in-relation}.
%\begin{definition}
%\label{def:append-first-version}
%Consider an object $A$, the relation $R$ and a time interval $J$. Then, the function append-first$\left(R,A,J \right)$ closes any right open version of the object $A$ if it exists and add the new version $\left(A, J \right)$.
%
%\begin{eqnarray}
%%\label{eq:close-current}
%\text{Append-first} \left(R, A, J \right) =\\
%\begin{cases}
%\nonumber
%\big \lbrace R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseL } \left(I, J\right) \right) \cup \left(A, J\right)\right \rbrace  \big \rbrace \\
%\nonumber
%\mbox{ if } I = \mbox{ First } \left(R, A \right) \wedge J < \mbox{ CloseL } \left(I, J \right)   \\
%\nonumber R , \text{ in any other case}
%\end{cases}
%\end{eqnarray}
%\end{definition}
\subsubsection{\label{subsubsec:modify}Modify}
This operation adds new information about an existing entity $A$ in the relation $R$. The modify operation does not remove any previous value of the object $A$. It closes the current version for the entity $A$ and adds a new version.

%Note that the modify operation is only applicable when the object $A$ is current in the relation $R$; $ \left(A, I \right) \in R, I = \left[a, +\infty \right]$.

\begin{align}
\label{eq:modify}
\MOD \left( R, \left(A, I \right)\right) =\\
%\begin{cases}
\nonumber
\mbox{ Close-current }\left(R, A, \mbox{Current}\left(R, A \right) \right) %\\
% \mbox{ if Current}\left(R ,A \right) \neq \emptyset \\
% R,  \mbox{ otherwise }
%\end{cases}
\end{align}

\subsubsection{\label{subsubsec:insert}Insert}
The user wants to store an object $A$ which is valid in the relation $R$ during the time interval specified by $I = \left(a, b \right)$.
%
%
%The interpretation of the insert operation is the following: The user wants to store in the database that an object $A$  is true for some period(s) of time $I$ in the given relation $R$. To indicate that the object $A$ is current in the relation, the special value Until Current, $\UC$  is used. 
There are two main cases when performing a create operation:
\begin{enumerate}
\item The object $A$ was never in the relation $R$: The object is added with the valid-time indicated by the crisp interval $I$. In the example, this correspond with the following insertion sentences:
\begin{verbatim}
Insert(3,'Dracula',
(15/3/2012, 30/3/2012));
Insert(4,'Frankenstein',
(25/3/2012, 4/4/2012));
Insert(5,'Harry Potter',
(18/3/2012, 2/4/2012));
\end{verbatim}

\item The object $A$ is in the relation $R$. Depending on the value of the time interval $I$, there are three possibilities:
	\begin{enumerate}
	\item Insert $\left(A, I\right)$ in the relation $R$. If the time interval $I$ does not overlap any other valid-time interval in the relation $R$. For example, consider that the book with ID 3 is loaned on 4/4/2012 but the book has not been returned yet. This is illustrated in Table \ref{tbl:library-sample}. The insert sentence is:
	      \begin{verbatim}
Insert(3,'Dracula',
(4/4/2012, UC));
	      \end{verbatim}

	\item Modify and close the current version of $A$ and insert $\left(A, I\right)$. For example, consider now that the book with ID 3 is loaned  the 24/4/2012. Here the problem is that the book with ID 3 was loaned on 4/4/2012, but not returned. If the book is already in the library for a new loan, then it is necessary to set the return date and add a new row with the new loan date. This is illustrated in Table \ref{tbl:library-sample-2}, the insert sentence is:
	    \begin{verbatim}
Insert(3,'Dracula',
(24/4/2012, UC));
	      \end{verbatim}
	\item Reject the insertion, if the time interval $I$ does overlap any existing valid-time interval for the object $A$ in the relation $R$. For example, consider now that the librarian wants to introduce a past loan for the book with ID 3. The book was loaned on 6/4/12/2012 and returned on 25/4/2012. As this interval does overlaps other time intervals, it is not possible that the book was loaned in that time interval.Therefore, the insertion is rejected. The insert sentence is:

	      \begin{verbatim}
Insert(3,'Dracula',
(6/4/12/2012,25/4/2012));
	    \end{verbatim}

@@

\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample-2}Sample library database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Title  & Start Loan & End Loan \\
\hline
3 & Dracula & 15/3/2012 & 30/3/2012 \\
4 & Frankenstein & 25/3/2012 & 4/4/2012 \\
5 & Harry Potter & 18/3/2012 & 2/4/2012\\
3 & Dracula & 4/4/2012 & 23/4/2012 \\
3 & Dracula & 24/4/2012 & UC \\
\hline\\
\end{tabular}
}

@@

\begin{align}
\label{eq:insert}
\INS \left(R ,\left(A, I\right)\right) =\\
\begin{cases}
\nonumber
R \cup \left(A, I \right) & \mbox{ if }  (A, I ) \not \in R  \vee \\
 &   \left(A, J\right) \in R \wedge \lnot \left(I \mbox{ Overlaps } J\right) \\
\mbox{modify(R,(A,I))} & \mbox{ otherwise }    %\mbox{ if } \exists J \left(A,J\right) \in R \wedge \mbox{Current(A,J)} \\ 
%R, &   
\end{cases} 	
\end{align}

\subsubsection{\label{subsubsec:del}Delete}
The delete operation logically removes a current entity $A$ which is valid in the relation $R$:
\begin{align}
\label{eq:delete}
\DEL \left(R ,A \right) =\\
\begin{cases}
\nonumber
R - \left(A, I \right), & \forall \left( A, I \right) \in R\\
R, & \mbox{ otherwise }  
\end{cases} 	
\end{align}

For example, consider that the librarian wants to delete the history for the book with ID = 3. The result of this operation is shown in Table \ref{tbl:library-sample-3}. The delete sentence is:

\begin{verbatim}
 Delete(3,'Dracula');
\end{verbatim}

\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample-3}Sample library database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Title  & Start Loan & End Loan \\
\hline
4 & Frankenstein & 25/3/2012 & 4/4/2012 \\
5 & Harry Potter IV & 18/3/2012 & 2/4/2012\\
\hline\\
\end{tabular}
}

\subsubsection{\label{subsubsec:revise}Revise}
This operation replaces the old values for the entity $A'= \left(a_1', \ldots, a_n' \right) $ with the new values specified by $A = \left(a_1, \ldots, a_n \right)$ in the relation $R$, at a given time interval $I$. This operation is defined when a correction on the values for the entity $A$ should be done.
\begin{align}
\REV\left(R, \left(A, I \right) \right)=\\
\begin{cases}
\nonumber
 R - \left(A', I \right) \cup \left(A, I \right) & \mbox{ if } \left(A', I \right) \in R  \\
   %: A = \left(a_1, \ldots, a_n \right),\\
%A'= \left(a_1', \ldots, a_n' \right)   \\   
R & \mbox{ otherwise }
\end{cases}
\end{align}

For example, consider now that the librarian wants to make a correction in the title for the book with ID = 5. The new title is `Harry Potter IV'. The result of this operation is illustrated in Table \ref{tbl:library-sample-3}.
 Then, the revise sentence is:

\begin{verbatim}
Revise(5,'Harry Potter IV', 
(18/3/2012, 2/4/2012));
\end{verbatim}
